Mining Pre-Exposure Prophylaxis Trends in Social Media

### 2016-05-20 draft/draft.tex
revise draft

Pre-Exposure Prophylaxis (PrEP) is a recently developed therapy for the prevention of Human Immunodeficiency Virus (HIV) transmission. The treatment is available as a pill named Truvada and is taken once a day by HIV negative individuals for as much as 99\% protection when exposed to HIV. While medically effective, PrEP suffers from high drug costs, HIV-related social stigma, and under-informed health providers and patients. Data mining of social media has proven effective in the monitoring of HIV in the United States, but no study has investigated PrEP using social media. This paper describes a data mining and machine learning strategy using natural language processing (NLP), that monitors Twitter social media data to identify PrEP discussion trends. Our results show that we can identify PrEP and HIV discussion dynamics over time, and PrEP-related tweets with positive and negative sentiment. These results can be used by public health professionals to monitor PrEP discussion trends, and identify strategies to improve HIV prevention.

@@

Word2vec is also used as a preprocessing step for additional analyses that start from pre-trained word-vectors. Doc2vec~\cite{le2014distributed} is a method that produces a document-vector for each document, and each document-level identifier. In the case of a tweet-corpus, document identifiers might include for example document ID, hashtags, and the user who created the document. Each of these document-level identifiers is embedded in a similarity space, allowing one to identify similar tweets, hastags, and users.

@@

In this paper we respond to the acknowledged need to harness large scale data in the battle against HIV~\cite{young2015big}. We use the above NLP techniques to extract PrEP-related semantic information from a Twitter corpus dataset. We identify critical PrEP related terms, users, hashtags and tweets. We identify PrEP discussion trends over time, and identify other topics that people who tweet about PrEP, also tweet about. Finally we train a sentiment classifier that automatically identifies PrEP-related tweets with positive and negative sentiment. Together these results, and these approaches, can be used by public health officials to identify trends in the national PrEP discussion and respond to public health issues as they arise.

\section{Results}
We sought to determine trends in HIV and PrEP discourse on twitter to inform and coordinate public health efforts aiming to promote PrEP adoption and adherence for at-risk individuals. We collected 624,569 tweets containing at least one of the following words 'HIV', 'AIDS', 'truvada', 'prophylaxis', 'imtesting', or 'PrEP' from Twitter's streaming API. The tweets were restricted to English language, and the collection dates spanned from the 47th week of 2015 to the 14th week of 2016. The tweets were cleaned of exotic characters. Before performing topic analysis, we excluded words that were mentioned fewer than 10 times, or more than 0.3 times the number of documents. We also performed Term Frequency Inverse Document Frequency (TF-IDF) normalization.

@@

The first analysis that we performed sought to identify certain keywords, hashtags, tweets and users, that were discussed in HIV and PrEP related contexts. Word2vec and the related method, doc2vec are unsupervised machine learning methods that have performed well at embedding natural language in a semantic vector space. In our analysis, word2vec allows us to determine semantically similar words to a query word, while doc2vec allows us to determine similar tweets, users and hashtags to a query hashtag.

@@

We investigated the blog article linked to tweet 708519265540907010\cite{prephurtingcause}. The article, authored by Steven Banning, an HIV positive individual, expresses concerns about how new HIV treatments, like PrEP are being adopted by the HIV community. Steven notes that treatments for HIV positive individuals, and treatments to prevent HIV such as PrEP have made HIV a much more treatable disease than it was in the 1980's. These medical advances may have led in part to some unintended medical and social consequences. Steven notes that the rise of drug-resistant HIV strains has increased, in part due to HIV patients not adhering to to consistent treatment plans. He also points out the recent social trend of ``bug chasers," individuals who are actively seeking to acquire and spread HIV strains. Steven Banning suggests that the widespread availability of effective HIV medications, may have desensitized healthy fear of the disease in some individuals.

@@

We identified two topics that showed relevant trends. In topic 5 we see that the keyword 'PrEP' increases over time, while 'prevention', 'drug' and 'risk' remain constant (Figure 3). The term 'pill', also from topic 5, declines overtime. These dynamics may indicate that PrEP discussion is becoming more prevalent. This growing trend would be consistent with the fact that while PrEP is still not widely known about among patients and healthcare providers, information about PrEP is slowly entering into national awareness. A study in New York City in 2011 indicated that only 36\% of high risk individuals were aware of PrEP~\cite{mehta2011awareness}.

Other HIV prevention related words such as 'pill', 'prevention' and 'drug' serve as negative controls. They are related to PrEP, but also used in other medical contexts. The fact that they are not increasing, shows us that the increase in PrEP discussion is PrEP-specific.

@@

We found at least one other DTM topic that showed interesting behavior. We observed that topic 4 captured several keywords related to World AIDS Day (Figure 4). We can see that ``WorldAIDSDay", and ``Can" peak in the 47th week of 2015 and then decline into 2016. This correlates well with the actual date of World AIDS Day, December 1st. Furthermore, while December 1st is World AIDS day, the whole month of December is AIDS Awareness Month. We can clearly see the words 'raise' and 'awareness' peak later and last longer than the word 'WorldAIDSDay' indicating that these words are correlated to the whole month of December. While our PrEP investigation isn't specifically interested in World AIDS Day, or AIDS Awareness Month, this observation serves as a control to validate our ability to accurately identify temporal events using DTM.

@@

% figure 5
\begin{figure*}
\centering
\includegraphics[height=2.5in, width=5in]{LDA_user_timelines}
\caption{Clustermap of users vs. topics for the top 500 users related to PrEP.}
\end{figure*}

@@

One of the top words from the HIV-related topic, topic 13, was the term ``stigma", the term ``endstigma" was also found in topic 17. Previous studies have shown a variety of stigmas associated and HIV have hampered prevention efforts~\cite{liu2014early}. Our observation of stigma related terms corroborates that there is some discussion of stigma in the context of HIV on Twitter. Public health professionals may be able use the prevalence of the term ``stigma" as a way to monitor the efficacy of efforts to end HIV related stigmas.

@@

Previously, we used analyses to summarize the whole Twitter corpus to produce high level trends. For our final analysis, we sought to get a deeper understanding of the data at the individual tweet level. Thus we trained a classifier to classify the sentiment of HIV and PrEP related tweets to be either positive or negative. This classifier would allow public health professionals to quickly identify positive and negative PrEP related tweets to guide HIV prevention efforts. We obtained a set of 1.6 million tweets with binary sentiment labels from Sanders Analytics~\cite{sentimentdata}. Then we trained a simple logistic regression classifier on 1.2 million paragraph-vectors from the sentiment dataset. We found that our classifier had an accuracy of 69\% using a portion of the sentiment tweets not used in training, as a validation set.

We chose to use a relatively simple classifier model (logistic regression) and stop training at 20 epochs of stochastic gradient decent over the corpus, because we wanted to prevent the possibility that we overtrained on our training data. This was especially important because our training and testing data, while both sets of tweets, were separate datasets. We then used this trained classifier to classify our PrEP related tweets into positive or negative sentiment labels. We identified the most positive, and most negative tweets, by log probability, on our full dataset, and on tweets that specifically mention either PrEP or Truvada. We provide the text from the top three positive and negative tweets from each of the three datasets (Table 3, Table 4).

@@

While the sentiment classification doesn't have perfect sentiment accuracy, we can see in general that the positive tweets are disseminating information about the efficacy and benefits of PrEP related preventions (Table 3). The positive tweets indicate that PrEP public health informational efforts have had some effect disseminating PrEP information on Twitter.

@@

Finally we see some contention over national political leadership in the effort to prevent HIV. While political discussions can often get heated, especially on Twitter, the contention can harm concerted efforts to provide consent governmental leadership in HIV prevention. Public health officials may consider stressing that unity in health policy at the federal level is important to protect the population from the spread of HIV.

The last tweet in the positive tweets table (Table 3), referencing the hashtag ``\#truvadawhore", links to a blog article written by Evan Peterson titled ``The Case for PrEP, or How I Learned to Stop Worrying and Love HIV-Positive Guys"~\cite{caseforprep}. Evan is an HIV-negative individual who adheres to a daily PrEP regimen to protect himself from HIV. Evan finds that for himself, the side effects and inconveniences associated with taking daily Truvada are worth the protection, noting that he also practices safe sex, and gets tested regularly. Evan addresses the hashtag \#truvadawhore, which originated with a Huffington Post article titled ``Truvada Whores?"~\cite{truvadawhore} which expressed concern toward the increased sexual risk taking of some Truvada patients. Evan notes that many HIV-negative individuals on Truvada like him who know that they may be exposed to HIV, are responsible and cautious about the risks they are taking. Individuals on Truvada may also be more informed than at risk people who do not use Truvada, because they are more educated and responsible with respect to their sexual health. Further large scale quantitative measurements will be needed to determine whether or not taking Truvada correlates with at risk behaviors.

@@

In this article we use a variety of NLP techniques to analyze a tweet corpus in order to monitor the national social media discussion for HIV prevention. We identified several notable trends. Firstly, we found that PrEP discussion is increasing on social media, Secondly we found that people who mention PrEP also talk about general health, STDs, stigma, and politics.

By picking out the most relevant tweets to PrEP, and also classifying tweets as positive or negative, we quickly identified a handful of specific tweets that highlight important parts of the online PrEP discussion. We identified two important arguments for and against PrEP. One, as explored by blog author Steven Banning hypothesizes that PrEP adoption could lead to unintended, and unhealthy risk seeking behavior. The contrary hypothesis, described by blog author Evan Peterson, suggests that PrEP adoption is a product of healthy, responsible at-risk individuals who are gaining HIV protection in an effective form. These opposing arguments were identified automatically using large scale social media data. Further investigation and data analysis, of social media data, or other health data may be able to quantify PrEP and Truvada's effectiveness in the ongoing effort to prevent HIV infection.

Future analyses may also take advantage of the rich Twitter API. An additional refinement of the analyses presented here may take a more active approach. Rather than simple passive data collection, we could build a twitter bot to interact with HIV and PrEP-tweeters, or conduct a social media based health survey. Finally, future HIV and PrEP research may benefit from the incorporation of multiple data sources, either multiple social media sources, multiple health information sources, or both.

@@

Acknowledgements (optional) go here.

>>>>

Pre-Exposure Prophylaxis (PrEP) is a recently developed therapy for the prevention of Human Immunodeficiency Virus (HIV) transmission. The treatment is available as a pill named Truvada and is taken once a day by HIV negative individuals for as much as 99\% protection when exposed to HIV. While medically effective, PrEP suffers from high drug costs, HIV-related social stigma, and under-informed health providers and patients. Data mining of social media has proven effective in the monitoring of HIV in the United States, but no study has investigated PrEP using social media. This paper describes a data mining and machine learning strategy using natural language processing (NLP), that monitors Twitter social media data to identify PrEP discussion trends. Our results show that we can identify PrEP and HIV discussion dynamics over time, and PrEP-related tweets with positive and negative sentiment. These results can be used by public health professionals to monitor PrEP discussion trends and identify strategies to improve HIV prevention.

@@

Word2vec is also used as a preprocessing step for additional analyses that start from pre-trained word-vectors. Doc2vec~\cite{le2014distributed} is one such method that uses pre-trained word-vectors to produce a document-vector for each document-level attribute. In the case of a tweet-corpus, document identifiers might include tweet ID, hashtags, and the user who created the tweet. Each of these document-level identifiers is embedded in a similarity space, allowing one to identify similar tweets, hastags, and users.

@@

In this paper we respond to the acknowledged need to harness large scale data in the battle against HIV~\cite{young2015big}. We use the above NLP techniques to extract PrEP-related semantic information from a Twitter corpus dataset. We identify critical PrEP related terms, users, hashtags and tweets. We identify PrEP discussion trends over time, and identify other topics that co-occur with who tweet about PrEP. Finally we train a sentiment classifier that automatically identifies PrEP-related tweets with positive and negative sentiment. Together these results, and these approaches, can be used by public health officials to identify trends in the national PrEP discussion and respond to public health issues as they arise.

\section{Results}
We sought to determine trends in HIV and PrEP discourse on Twitter to inform and coordinate public health efforts aiming to promote PrEP adoption and adherence for at-risk individuals. We collected 624,569 tweets containing at least one of the following words 'HIV', 'AIDS', 'truvada', 'prophylaxis', 'imtesting', or 'PrEP' from Twitter's streaming API. The tweets were restricted to English language, and the collection dates spanned from the 47th week of 2015 to the 14th week of 2016. The tweets were cleaned of exotic characters. Before performing topic analysis, we excluded words that were mentioned fewer than 10 times, or more than 0.3 times the number of documents. We also performed Term Frequency Inverse Document Frequency (TF-IDF) normalization.

@@

The first analysis that we performed sought to identify certain keywords, hashtags, tweets and users, that were discussed in HIV and PrEP related contexts. Word2vec and the related method, doc2vec, are unsupervised machine learning methods that have performed well at embedding natural language in a semantic vector space. In our analysis, word2vec allows us to determine semantically similar words to a query word, while doc2vec allows us to determine similar tweets, users and hashtags to a query hashtag.

@@

We investigated the blog article linked to tweet 708519265540907010\cite{prephurtingcause}. The article, authored by Steven Banning, an HIV positive individual, expresses concerns about how new HIV treatments, like PrEP are being adopted by the HIV community. Steven notes that treatments for HIV positive individuals, and treatments to prevent HIV such as PrEP, have made HIV a much more treatable disease than it was in the 1980's. These medical advances may have led in part to some unintended medical and social consequences. Steven notes that the rise of drug-resistant HIV strains has increased, in part due to HIV patients not adhering to to consistent treatment plans. He also points out the recent social trend of ``bug chasers," individuals who are actively seeking to acquire and spread HIV strains. Steven Banning suggests that the widespread availability of effective HIV medications, may have desensitized a healthy sense of fear of the disease in some individuals.

@@

We identified two topics that showed relevant trends. In topic 5 we see that the keyword `PrEP' increases over time, while `prevention', `drug' and `risk' remain constant (Figure 3). The term `pill', also from topic 5, declines overtime. These dynamics may indicate that PrEP discussion is becoming more prevalent. This growing trend would be consistent with the fact that while PrEP is still not widely known about among patients and healthcare providers, information about PrEP is slowly entering into national awareness. A study in New York City in 2011 indicated that only 36\% of high risk individuals were aware of PrEP~\cite{mehta2011awareness}.

Other HIV prevention related words such as `pill', `prevention' and `drug' serve as negative controls. They are related to PrEP, but also used in other medical contexts. The fact that they are not increasing, shows us that the increase in PrEP discussion is PrEP-specific.

@@

We found at least one other DTM topic that showed interesting behavior. We observed that topic 4 captured several keywords related to World AIDS Day (Figure 4). We can see that ``WorldAIDSDay", and ``Can" peak in the 47th week of 2015 and then decline into 2016. This correlates well with the actual date of World AIDS Day, December 1st. Furthermore, while December 1st is World AIDS day, the whole month of December is AIDS Awareness Month. We can clearly see the words `raise' and `awareness' peak later and last longer than the word `WorldAIDSDay' indicating that these words are correlated to the whole month of December. While our PrEP investigation isn't specifically interested in World AIDS Day, or AIDS Awareness Month, this observation serves as a control to validate our ability to accurately identify temporal events using DTM.

@@

% figure 5
\begin{figure*}
\centering
\includegraphics[height=2.5in, width=5in]{LDA_user_timelines}
\caption{LDA topic modeling for the top 500 users related to PrEP.}
\end{figure*}

@@

One of the top words from the HIV-related topic, topic 13, was the term ``stigma", the term ``endstigma" was also found in topic 17. Previous studies have shown a variety of stigmas associated and HIV have hampered prevention efforts~\cite{liu2014early}. Our observation of stigma related terms corroborates that there is some discussion of stigma in the context of HIV on Twitter. Public health professionals may be able use the prevalence of the term ``endstigma" as a way to monitor the efficacy of efforts to end HIV related stigmas.

@@

Previously, we used analyses to summarize the whole Twitter corpus to extract high level trends. For our final analysis, we sought to get a deeper understanding of the data at the individual tweet level. Thus we trained a classifier to classify the sentiment of HIV and PrEP related tweets to be either positive or negative. This classifier would allow public health professionals to quickly identify positive and negative PrEP related tweets to guide HIV prevention efforts. We obtained a set of 1.6 million tweets with binary sentiment labels from Sanders Analytics~\cite{sentimentdata}. Then we trained a simple logistic regression classifier on 1.2 million paragraph-vectors from the sentiment dataset. We found that our classifier had an accuracy of 69\% using a portion of the sentiment tweets not used in training, as a validation set.

We chose to use a relatively simple classifier model (logistic regression) and stop training at 20 epochs of stochastic gradient decent over the corpus, because we wanted to prevent the possibility that we overtrained on our training data. This was especially important because our training and testing data, while both sets of tweets, were separate datasets. We used this trained classifier to classify our PrEP related tweets into positive or negative sentiment labels. We identified the most positive, and the most negative tweets, by log probability, on our full dataset, and positive and negative tweets that specifically mention either PrEP or Truvada. We provide the text from the top three positive and negative tweets from each of the three datasets (Table 3, Table 4).

@@

While the sentiment classification doesn't have perfect sentiment accuracy, we can see in general that the positive tweets are disseminating information about the efficacy and benefits of PrEP related preventions (Table 3). The positive tweets indicate that PrEP public health informational efforts have had some effect disseminating useful PrEP information on Twitter.

@@

Finally we see some contention over national political leadership in the effort to prevent HIV. While political discussions can often get heated, especially on Twitter, the contention can harm concerted efforts to provide consistent governmental leadership in HIV prevention. Public health officials may consider stressing that unity in health policy at the federal level is important to protect the population from the spread of HIV.

The last tweet in the positive tweets table (Table 3), referencing the hashtag ``\#truvadawhore", links to a blog article written by Evan Peterson titled ``The Case for PrEP, or How I Learned to Stop Worrying and Love HIV-Positive Guys"~\cite{caseforprep}. Evan is an HIV-negative individual who adheres to a daily PrEP regimen to protect himself from HIV. Evan finds that for himself, the side effects and inconveniences associated with taking daily Truvada are worth the protection, noting that he also practices safe sex, and gets tested regularly. Evan addresses the hashtag \#truvadawhore, which originated with a Huffington Post article titled ``Truvada Whores?"~\cite{truvadawhore} which expressed concern toward the increased sexual risk taking of some Truvada patients. Evan provides a rebuttal, noting that many HIV-negative individuals on Truvada, including himself and people he knows, are responsible and cautious about the risks they are taking. He explains that taking daily Truvada may actually be an indication that individuals are more educated and responsible with respect to their sexual health. Further large scale quantitative measurements, including large scale medical surveys, will be needed to determine whether or not taking Truvada correlates positively or negatively with risk-taking behaviors.

@@

In this article we use a variety of NLP techniques to analyze a tweet corpus in order to monitor the national social media discussion for HIV prevention. We identified several notable trends. We found that PrEP discussion activity is increasing on social media, and also that people who mention PrEP also talk about general health, STDs, stigma, and politics.

By picking out the most relevant tweets to PrEP, and also classifying tweets as positive or negative, we quickly identified a handful of specific tweets that highlight important parts of the online PrEP discussion. We identified two important arguments for and against PrEP. One, as explored by blog author Steven Banning hypothesizes that PrEP adoption could lead to unintended, and unhealthy risk seeking behavior. The contrary hypothesis, described by blog author Evan Peterson, suggests that PrEP adoption is a product of healthy, responsible at-risk individuals who are gaining an effective form of HIV protection. These opposing arguments were identified automatically using large scale social media data. Further investigation and data analysis, of social media data together with other health data may be able to quantify PrEP and Truvada's effectiveness in the ongoing effort to prevent HIV infection.

Future analyses may also take further advantage of the rich Twitter API. Rather than simple passive data collection, we could build a twitter bot to interact with HIV and PrEP-tweeters, to conduct a large scale social-media-based health survey.

@@

We would like to acknowledge the Twitter streaming API, which we used to collect HIV and PrEP related tweets over a 30 week period from late 2015 to early 2016. Also, we used the gensim python library for our working implementations of word2vec, doc2vec, LDA, and DTM. A repository containing the scripts used in this analysis (but no data) is available at https://github.com/quinngroup/sm\_w2v.
